---
title: "Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r sessione_info, include=FALSE}
sessioninfo::session_info()
```

# Genetic Algorithm Optimization Vignette

This vignette provides a step-by-step explanation of a genetic algorithm 
optimization process using the BioGA package.

## Chapter 1: Initialization

```{r}
# Load the BioGA package
library(BioGA)

# Example genomic data
genomic_data <- matrix(rnorm(50), nrow = 10, ncol = 10)

# Initialize population
population <- initialize_population_cpp(genomic_data, population_size = 5)

# Initialize fitness history
fitness_history <- list()

# Initialize time progress
start_time <- Sys.time()
```

## Chapter 2: Genetic Algorithm Optimization

```{r}
# Run genetic algorithm optimization
generation <- 0
while (TRUE) {
    generation <- generation + 1

    # Evaluate fitness
    fitness <- evaluate_fitness_cpp(genomic_data, population)
    fitness_history[[generation]] <- fitness

    # Check termination condition
    if (generation == 6) {
        break
    }

    # Selection
    selected_parents <- selection_cpp(population, fitness,
        num_parents = 2
    )

    # Crossover and Mutation
    offspring <- crossover_cpp(selected_parents, offspring_size = 2)
    # (no mutation in this example)
    mutated_offspring <- mutation_cpp(offspring, mutation_rate = 0)

    # Replacement
    population <- replacement_cpp(population, mutated_offspring,
        num_to_replace = 1
    )

    # Calculate time progress
    elapsed_time <- difftime(Sys.time(), start_time, units = "secs")

    # Print time progress
    cat(
        "\rGeneration:", generation, "- Elapsed Time:",
        format(elapsed_time, units = "secs"), "     "
    )
}
```

## Chapter 3: Results Visualization

```{r}
# Plot fitness change over generations
plot_fitness_history(fitness_history)
```

This vignette demonstrates how to implement a genetic algorithm optimization
process using the BioGA package, from initialization to results 
visualization.



<details>

<summary>**Session Info**</summary>

```{r sessioninfo}
sessioninfo::session_info()
```

</details>
